Phenotype and risk assesment of insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry

### 2019-10-14 analysis/paper/paper.Rmd
corrections to figures and results round 3

@@

## VIA is often associated with cardiovascular symptoms

@@

## Absence of skin symptoms is associated with more severe episodes of VIA

Patients with concomitant mastocytosis had significantly more often anaphylaxis without skin symptoms (namely urticaria and flushing) when compared to patients without diagnosed mastocytosis (`r (skin_symptoms$masto_tab$prop[c(2,1)]*100) %>% round(1) %>% paste0(collapse="% vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`). We decided to further investigate non-mastocytic patients. In these patients urticaria or flushing as symptoms of anaphylaxis were present less often in VIA than anaphylaxis triggered by other elicitors (`r (skin_symptoms$tab$prop[2]*100) %>% round(1)`% vs `r (skin_symptoms$tab$prop[4]*100) %>% round(1)`% respectively, `r pval(skin_symptoms$chisq$p.value)`).  Morover, in patients without concomitant mastocytosis who were also lacking skin symptoms, anaphylaxis was significantly more frequently severe (`r (skin_symptoms$severe_tab$prop[c(2,6)]*100) %>% round(1) %>% paste0(collapse="% in VIA vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`% in non-VIA). We performed a factorial logistic regression modelling and confirmed the significant interaction effect between the presence of skin symptoms and the trigger (eiter VIA or non-VIA) on the severity of anaphylaxis (`r pval(skin_symptoms$severity_fit_skin_symptoms %>% coef %>% .[4,4])`). Therefore, non-mastocytic patients presenting without urticara or flushing tended to have more severe anaphylaxis when it was triggered by insects, but not other elicitors of anaphylaxis (Fig. \@ref(fig:figsymptoms)C).

Following this observation, we decided to investigate the association of skin symptoms with the tryptase levels in patients who did not have a diagnosis of mastocytosis. For this comparison, we removed the cases with BST above 11.5 ng/ml, potentially indicating non-diagnosed mastocytosis. Similarly, tryptase levels correlated with the severity of anaphylaxis, were higher in VIA patients, and the effect was prominently visible for VIA patients (`r skin_symptoms$fit_tryp[[1]][4,5] %>% pval()`). We did not observe such an interaction in the non-VIA group (Fig. \@ref(fig:figsymptoms)D).

## Severity 
Severe reactions of VIA were more prevalent in older patients in comparison to patients below 22 years, and in VIA cases vs. other elicitors (Fig. S\@ref(fig:sev)D). There were no differences in severity of reactions elicited by yellow-jackets and other insect species.

@@

## Increased baseline serum tryptase and concomitant cardiovascular conditions increase the risk of severe VIA.

The factor most prominently associated with an increased risk of severe anaphylaxis was mastocytosis, and there were no differences in elicitor groups (Fig. \@ref(fig:figcofactors)). However, mastocytosis increased the risk of cardiac arrest in patients undergoing VIA significantly more than in patients undergoing anaphylaxis due to other triggers (Fig. S\@ref(fig:supplement)A).

The risk of severe anaphylaxis in patients concomitantly using ACE-I (as well as beta-blockers) could not be independently measured due to coexisting cardiovascular pathologies. ACE-I use was, however, more often associated with cardiac arrests in all anaphylaxis cases (`r ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(prop) %>% pull() %>% rev() %>% compare()`, `r pval(ace_tab %>% dplyr::select(-prop) %>% spread(q_114_cardiac_arrest,n) %>% {chisq.test(.[,2:3])$p.value})`) and there were no differences between VIA and non-VIA  (Fig. \@ref(fig:figcofactors)C). 
Beta-blocker use was generally associated with higher severity of anaphylaxis and with the onset of cardiovascular symptoms (cardiac arrest, chest pain), but showed no difference between groups `r summary(beta_fit)[["coefficients"]][3,4] %>% pval()`). Surprisingly, arrhythmia was more frequently reported in patients with VIA and concomitant beta-blockers (Fig. \@ref(fig:figcofactors)C).

Baseline serum tryptase (BST) correlated with the severity of reactions (Ring and Messmer scale). Elevated BST was more prominently increasing the risk of severe anaphylaxis in VIA than in non-VIA (Fig. \@ref(fig:figsymptoms)D and Fig. \@ref(fig:figcofactors)B).
Cases with cardiac arrest were associated with an increased tryptase above 8 ng/ml. This proportion was higher in VIA when compared to other elicitors. Loss of consciousness was a symptom associated with increased tryptase levels but only in patients with VIA (Fig. \@ref(fig:figcofactors)C). Based on the severity and symptom profile, we defined a tryptase cut-off value of 8 ng/ml (Fig. \@ref(fig:figcofactors)B) instead of the currently used 11.5 ng/ml.

The Hymenoptera species responsible for triggering VIA was independent from  the severity of the reaction. Patients matched according to sex and age had similar severity of a reaction to known eliciting insects (p = 0.4128). 
<!-- This data, however, is not comparable to the refractory anaphylaxis cases as refractory anaphylaxis seems to be a distinct phenotype.  -->

@@

## Interplay of symptoms and medication

@@

We discovered clusters of symptoms and therapy modes in VIA patients (association measured using phi). Cardiovascular symptoms (cardiac arrest, hypotension, loss of consciousness) and urticaria, were treated more similarly than respiratory or gastrointestinal symptoms (Fig. \@ref(fig:adrenalineuse)B). The treatment of these symptoms consisted of adrenaline autoinjector (AAI) use, i.v. adrenaline in multiple doses, 100% oxygen inhalation, an initial dose of antihistamines, and inhaled β-2 agonists. 
Corticosteroids, i.v. volume replacement and i.v. β-2 agonists formed another therapy mode.

The most intriguing difference in the therapy of VIA vs. other forms of anaphylaxis were regarding the frequency of inhaled beta2-agonists and antihistamines (Fig. \@ref(fig:adrenalineuse)B).

@@

## Hymenoptera anaphylaxis is a seasonal disease.

@@

In this study, we identified the prolonged seasonality of VIA, its clinical symptom-profile, and treatment patterns. The data unravel phenotypes of VIA, which may contribute to the development of tools incorporating both clinical <!--- and biomarker--->data for predicting the severity of future episodes of anaphylaxis.

@@

# Acknowledgements
We thank all patients, parents, and their children for their support in providing data on the occurrence of anaphylaxis for this study. We thank the study personnel for patients counseling and data entry, and we thank the members of The European Anaphylaxis Registry in detail:

@@

```{r figsymptoms, fig.cap="Symptoms of insect venom anaphylaxis (VIA) compared to other elicitors. A: Proportional presentation of specific reaction symptoms. B: High-level overview of involved organ systems and selected cofactors in the form of a radar plot. C: Interaction effect of VIA and lack of skin symptoms (urticaria or flushing) on the severity of anaphylaxis. D: Baseline Serum Tryptase (BST) increases with the severity of anaphylaxis a lot more prominently in VIA.", fig.width=8.5, fig.height = 6.2}
fig_symptoms2
```

@@

>>>>

Based on the severity and symptom profile, we defined a sub-elevated baseline serum tryptase (BST) cut-off value of above 8 ng/ml (Fig. \@ref(fig:figcofactors)B) as being more sensitive indicator compared to the currently used 11.5 ng/ml.

@@

## VIA is associated with cardiovascular symptoms

@@

Younger patients (under 22) presented even more prominent differences in hypotension symptoms and more frequently reported gastrointestinal symptoms (e.g. vomiting) **when the reaction was not triggered by insect venom**.(Fig. \@ref(fig:figsymptoms)C-E)

## Absence of skin symptoms during anaphylaxis is associated with more severe episodes of VIA

We found that `r (skin_symptoms$masto_tab$prop[c(2)]*100) %>% round(1)`% of patients with concomitant mastocytosis had anaphylaxis without skin symptoms (i.e. urticaria and flushing) which was significantly more compared to patients without diagnosed mastocytosis (`r (skin_symptoms$masto_tab$prop[c(1)]*100) %>% round(1)`%, `r pval(skin_symptoms$masto_chisq$p.value)`) and this observarion was most prominently seen in VIA (Fig. \@ref(fig:figskinsymptoms)).

Similarily, in non-mastocytic patients undergoing VIA, urticaria or flushing as symptoms of anaphylaxis were present less often than if anaphylaxis was triggered by other elicitors (`r (skin_symptoms$tab$prop[2]*100) %>% round(1)`% vs `r (skin_symptoms$tab$prop[4]*100) %>% round(1)`% respectively, `r pval(skin_symptoms$chisq$p.value)`). Moreover, in this specific subgroup of patients (i.e. non-mastocytic and at the same time lacking urticaria or flushing) VIA was significantly more frequently severe (`r (skin_symptoms$severe_tab$prop[c(6,2)]*100) %>% round(1) %>% paste0(collapse="% in VIA vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`% in non-VIA).

After performing a factorial logistic regression modelling we confirmed the significant interaction effect between the presence of skin symptoms and the trigger (VIA) on the severity of anaphylaxis (`r pval(skin_symptoms$severity_fit_skin_symptoms %>% coef %>% .[4,4])`). To recapitulate, non-mastocytic patients presenting without urticaria or flushing tended to have more severe anaphylaxis when it was triggered by insects, but not other elicitors of anaphylaxis (Fig. \@ref(fig:figskinsymptoms)C, and Tab. S \@ref(tab:factoriallogit)).

## Absence of skin symptoms correlates with BST levels and increases the risk of severe anaphylaxis specifically in IVA

Following the above-mentioned observation, we investigated the association of skin symptoms with the tryptase levels in non-masctoscytic patients. For this model, we excluded the cases with known mastocytosis and with BST above 11.5 ng/ml, potentially indicating non-diagnosed mastocytosis. Similarily, 1) tryptase levels correlated with the severity of anaphylaxis, 2) were higher in VIA patients, and 3) the effect was more prominently visible in VIA (`r skin_symptoms$fit_tryp[[1]][4,5] %>% pval()`). We did not observe this effect in the non-VIA group (Fig. \@ref(fig:figskinsymptoms)D).

## Sub-elevated baseline serum tryptase and concomitant cardiovascular conditions increase the risk of severe VIA.

We evaluated potential cofactors increasing the risk of severe IVA. The factor most prominently associated with an increased risk of severe anaphylaxis was mastocytosis (Fig. \@ref(fig:figcofactors)), and it increased the risk of: 1) cardiac arrest and 2) loss of consciousness in patients undergoing VIA significantly more than in patients undergoing anaphylaxis due to other triggers (Fig. S\@ref(fig:supplement)A and Fig. \@ref(fig:figcofactors)C).

In line with the above, BST levels also correlated with the severity of anaphylaxis (on the Ring and Messmer scale) and, most importantly, sub-elevated BST was more prominently increasing the risk of severe anaphylaxis in VIA than in non-VIA (Fig. \@ref(fig:figsymptoms)D and Fig. \@ref(fig:figcofactors)B).

## Other cofactors of severe reactions (potentially could be shifted to th supplement)

Severe reactions of VIA were more prevalent in patients above 22 years of age, and in VIA cases vs. other elicitors (Fig. S\@ref(fig:sev)D). There were no differences in severity of reactions elicited by yellow-jackets and other 
insect species (p = 0.4128). .

The effect of using ACE-I (as well as beta-blockers) on the risk of severe anaphylaxis could not be independently measured from the coexisting cardiovascular pathologies. ACE-I use was, however, more often associated with cardiac arrests in all anaphylaxis cases (`r ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(prop) %>% pull() %>% rev() %>% compare()`, `r pval(ace_tab %>% dplyr::select(-prop) %>% spread(q_114_cardiac_arrest,n) %>% {chisq.test(.[,2:3])$p.value})`) and there were no differences between VIA and non-VIA  (Fig. \@ref(fig:figcofactors)C). 
Beta-blocker use was generally associated with higher severity of anaphylaxis and with the onset of cardiovascular symptoms (cardiac arrest, chest pain), but showed no difference between groups `r summary(beta_fit)[["coefficients"]][3,4] %>% pval()`). Surprisingly, arrhythmia was more frequently reported in patients with VIA and concomitant beta-blockers (Fig. \@ref(fig:figcofactors)C).

@@

<!-- This data, however, is not comparable to the refractory anaphylaxis cases as refractory anaphylaxis seems to be a distinct phenotype.  -->

@@

## Therapy: VIA recived adrenaline less often than non-VIA

@@

We discovered clusters of symptoms and therapy modes in VIA patients (association measured using phi coefficient). Cardiovascular symptoms (cardiac arrest, hypotension, loss of consciousness) and urticaria, were treated more similarly than respiratory or gastrointestinal symptoms (Fig. \@ref(fig:adrenalineuse)B). The treatment of these symptoms consisted of adrenaline autoinjector (AAI) use, i.v. adrenaline in multiple doses, 100% oxygen inhalation, an initial dose of antihistamines, and inhaled β-2 agonists. 
Corticosteroids, i.v. volume replacement and i.v. β-2 agonists formed another therapy mode.

The most prominent difference in the therapy of VIA vs. other forms of anaphylaxis were regarding the frequency of inhaled beta2-agonists and antihistamines (Fig. \@ref(fig:adrenalineuse)B).

@@

## Hymenoptera anaphylaxis is a seasonal disease (may be moved to supplement).

@@

In this study, we identified distinctive clinical symptom-profile, and treatment patterns of venom induced anaphlaxis. The data unravel phenotypes of VIA, which may contribute to the development of tools incorporating clinical <!--- and biomarker--->data for predicting the severity of future episodes of anaphylaxis.

@@

# Acknowledgements
We thank all patients, parents, and their children for their support in providing data on the occurrence of anaphylaxis for this study. We thank Dr. rer. biol. Kristin Franke. We thank the study personnel for patients counseling and data entry, and we thank the members of The European Anaphylaxis Registry in detail:

@@

```{r figsymptoms, fig.cap="Symptoms of insect venom anaphylaxis (VIA) compared to other elicitors. A: Proportional presentation of specific reaction symptoms. B: High-level overview of involved organ systems and selected cofactors in the form of a radar plot. C: Interaction effect of VIA and lack of skin symptoms (urticaria or flushing) on the severity of anaphylaxis. D: Baseline Serum Tryptase (BST) increases with the severity of anaphylaxis a lot more prominently in VIA.", fig.width=8.5, fig.height = 6.2}
fig_symptoms_3
```

```{r figskinsymptoms, fig.cap="Lack of skin symptoms during anaphylaxis is associated with more severe IVA."}

@@

```{r factoriallogit, results="asis"}
stargazer::stargazer(fit_skin_symptoms, type = "html",out = "stargazer_tab.doc")
